---
title: "How to Simulate Water Balance with R-Abimo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Simulate Water Balance with R-Abimo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(kwb.rabimo)
```

# Background

The R package kwb.rabimo is a result of the research project 
[AMAREX](https://amarex-projekt.de/de) (an acronym for the German translation of 
"adaptation of stormwater management to extreme events"), funded by the former 
Federal Ministry of Education and Research ([Bundesministerium für Bildung und 
Forschung -- BMBF](https://www.bmbf.de/DE/Home/home_node.html)).

In this project, the Water Balance Model ABIMO ()

[Wasserhaushaltsmodell Berlin ABIMO 3.2](https://github.com/umweltatlas/abimo)


code in this package has been transferred from the C++
code of ABIMO 3.3: Water Balance Model for Urban Areas
(https://github.com/KWB-R/abimo/).

## Installation

```r
# Install package "remotes" from CRAN
install.packages("remotes", repos = "https://cloud.r-project.org")

# Install package "kwb.rabimo" (latest "release") from GitHub
remotes::install_github("KWB-R/kwb.rabimo")

# Install package "kwb.rabimo" (development version) from GitHub
remotes::install_github("KWB-R/kwb.rabimo@dev")
```

## Basic Usage

### Provide input data and configuration

Compared to the original C++ version of Abimo we have modified the structures
of input data, output data and configuration.

For the German city of Berlin, we provide data in the new structures in the 
package:

```r
# Load Berlin data in the original Abimo format
abimo_inputs <- kwb.rabimo::rabimo_inputs_2025
```

### Run R-Abimo for the status quo

```r
# Run R-Abimo, the R-implementation of Abimo
rabimo_result <- kwb.rabimo::run_rabimo(
  data = abimo_inputs$data, 
  config = abimo_inputs$config
)

# Have a look at the first lines of the result data frame
head(rabimo_result)
```

### Run R-Abimo for a natural state scenario

```r
rabimo_result_natural <- kwb.rabimo::run_rabimo(
  data = kwb.rabimo::data_to_natural(abimo_inputs$data), 
  config = new_inputs$config
)
```

### Calculate "Delta-W"

For the first ten blocks, calculate the deviation from the natural state:

```r
kwb.rabimo::calculate_delta_w(
  urban = rabimo_result[1:10, ],
  natural = rabimo_result_natural
)
```

## Documentation

Release: [https://kwb-r.github.io/kwb.rabimo](https://kwb-r.github.io/kwb.rabimo)

Development: [https://kwb-r.github.io/kwb.rabimo/dev](https://kwb-r.github.io/kwb.rabimo/dev)
